---
title: The OES Guide to Making and Evaluating Matched Designs (DRAFT!!)
author: Jake Bowers and Ben Jaques-Leslie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
bibliography: big.bib
execute:
  echo: true
  output: true
  message: false
  warning: false
  freeze: auto
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    code-fold: false
    fig-caption: true
    code-links:
      - text: Github
        icon: github
        repo: https://github.com/bowers-illinois-edu/OES-Matching-Guide
---

```{r loadlibraries}
#| code-fold: true
#| code-summary: "Code to load libraries and general setup"
if(!require(renv)){
    install.packages("renv",method="wget")
}
library(tidyverse)
library(RItools)
library(optmatch)
library(designmatch)
library(highs)
library(coin)
library(conflicted)
library(estimatr)
library(senstrat)
library(sensemakr)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)
conflicts_prefer(optmatch::strata)
options(digits=4,scipen=8)
```

# Overview

We create matched research designs to address alternative explanations. Say we
have a policy intervention, $Z$, that we code as $Z_i=0$ for people, $i$, who
did not experience the intervention and $Z_i=1$ for those who did experience
the intervention, and an outcome that we observe for each person, $Y_i$. We
would like to know whether the outcome among people who experienced the
intervention  ($Y_i|Z_i=1$) is different from the outcome among those who did
not experience the intervention  ($Y_i|Z_i=0$) **because of the intervention
rather than some other factor, $x_i$**. A matched design will only compare
$Y_i|Z_i=1$ with $Y_i|Z_i=0$ **within strata of people who are similar on their
values of $x_i$ in an effort to set aside $x_i$ as an explanation of the
difference in outcomes.**

For example, say the intervention is a job training program and the outcome is
income and we see that people who took the job training program have higher
incomes than people who did not. One explanation for the difference could be
that the job training program helped people earn more. An alternative
explanation might be that that older people took the job training program and,
because they have more workforce experience tend to earn higher wages than
younger people and were also more likely to sign up for the job training
program. We can address this alternative explanation by creating strata of
people who have the same age and only comparing incomes between those who took
the training program and those who did not among people of the same age. For
example, we might write the difference in average income between job training
alumni and those who did not take job training among people who are 18 years
old as $\bar{Y}=(\bar{Y}|Z=1,X=18)-(\bar{Y}|Z=0,X=18)$. Within that strata of
18 year olds, there are no differences in age. So, comparisons in income
between those with the job training and without cannot be interepreted as
arising from differences in age (unless age differences in month or week
matter). If age differences tend to only matter if the differences in age are a
matter of years, then we can create a research design that **adjusts for this
alternative explanation** by restricting our comparisons to people who are the
same age: creating strata of 18 year olds, 19 year olds, etc $\ldots$.

In this memo, we discuss generalizations of this basic approach: from exactly
matching on a single variable to exclude alternative explanations based on that
variable alone, to matching on a number of variables all at once in the hopes
that we might set aside a number of alternative explanations all together.

Although we eventually will want to estimate values for and test hypotheses
about causal effects, this creation of a matched design is a research design
step because it can be done without knowledge of the values for the outcome
variables.^[It is useful to know which units have missing data on the outcomes,
but it is not necessary to know the values of the outcomes.] This separation of
design from outcome analysis allows us to use the data to make design decisions
without worrying about harming the false positive rates of our tests. Those who
wish to follow up on the general ideas presented here might like to see
@rosenbaum2020modern, @rosenbaum2020book.

This memo also is a code tutorial, showing examples throughout. It is currently
only using R because that is the language where those who have been developing
modern approaches to the creation of matched designs are working.

## Roadmap

This guide proceeds in four broad stages.

**Stage 1: The simplest case.** We begin with a matched design that has only two
strata and one covariate. This simple case introduces the core ideas of
stratification, within-strata estimation, and weighted combination of
strata-specific effects --- ideas that carry through everything that follows.

**Stage 2: Creating a matched design.** We then show how to generalize from two
strata to hundreds, and from one variable to many, using optimal matching
algorithms. We cover two R packages: `optmatch`, which uses network flow
optimization, and `designmatch`, which uses integer programming. Both support
*bipartite matching* (one treated group vs. one control group) and we also show
*non-bipartite matching* for continuous or multi-level treatments. We demonstrate
the key tools for controlling how the algorithm works: distance matrices
(Mahalanobis and propensity score), calipers (which forbid poor matches),
and exact matching constraints (which force perfect balance on key variables).

**Stage 3: Evaluating and improving the design.** A matched design is not
accepted or rejected in a single step --- it is iterated. We show two
complementary ways to assess a design: inspecting within-set differences on
substantively important variables, and comparing the overall design to the
covariate balance we would expect from a randomized experiment using the omnibus
test in `balanceTest`. We explain in detail what that test is doing so that its
output can be interpreted and acted upon.

**Stage 4: From design to inference.** Once satisfied with the design, we show
how to estimate average treatment effects and test hypotheses using the matched
sets as strata, following the same logic as a block-randomized experiment. We
close with sensitivity analysis --- tools for reasoning about how large an
unobserved confounder would need to be to change our conclusions.

# Create Data

To make this guide portable, we are making up a fake dataset with covariates
that help determine the probability of treatment.

```{r makedata}
#| echo: true
#| output: false
#| code-fold: true
N <- 1000
set.seed(12345)
dat <- data.frame(id = 1:N)
dat <- dat %>% mutate(
  sso = sample(1:10, size = N, replace = TRUE),
  age = rnorm(n = N, mean = 45, sd = 10),
  sex = factor(sample(c("M", "F"), size = N, replace = TRUE)),
  race = factor(sample(c("A", "B", "W"), size = N, replace = TRUE, prob = c(.05, .15, .8))),
  u = rnorm(n = N, mean = sso, sd = exp(sso)),
  demog_latent0 = .5 * sd(u) * age + .2 * sd(u) * as.numeric(sex) + .2 * sd(u) * as.numeric(race) + u,
  demog = (demog_latent0 - min(demog_latent0)) / (max(demog_latent0) - min(demog_latent0)),
  edu = factor(sample(letters[1:8], size = N, prob = runif(n = 8, min = .1, max = pmax(.2, demog)), replace = TRUE)),
  grade = sample(1:6, size = N, replace = TRUE),
  pay = abs(rnorm(n = N, mean = (demog + .01) * 50000 + 50000, sd = 50000)),
  tenure = rnorm(n = N, mean = 10, sd = 3),
  perf = runif(n = N, min = 1, max = 10),
  preY0 = abs(.2 * demog + rnorm(N)),
  preY = (preY0 - min(preY0)) / (max(preY0) - min(preY0)),
  trtcont = rnorm(n = N, mean = demog + preY, sd = sd(demog + preY)),
  ##  trtbin = rbinom(n = N, size = 1, prob = pmin(.95,demog+preY)),
  trtbin = as.numeric(trtcont >= quantile(trtcont, 2 / 3)),
  postY = .2 * sd(preY) * trtbin + preY + rnorm(N)
)

## optmatch will need row names
row.names(dat) <- dat$id
## Just verifying that trtbin is related to demog and preY
## table(dat$trtbin, exclude = c())
## with(dat, boxplot(demog ~ trtbin))
## with(dat, boxplot(preY ~ trtbin))

## Check the relationship between demog and its components. Not really using demog directly below.
## The idea is to ensure some relationships among the covariates using demog.
lm1 <- lm(demog ~ sso + age + sex + race + edu, data = dat)
#  summary(lm1)
lm2 <- lm(preY ~ demog, data = dat)
# summary(lm2)

# cor(dat[, c("age", "demog", "pay", "preY")])
```


Our goal is to assess the effect of some observed intervention (that we call
`trtbin` here for the binary version or `trtcont`) on some outcome (that we
call `postY`). We worry that a naive description of the relationship between
intervention and outcome would not provide useful policy guidance because any
such simple two-way relationship might contain other relationships (say, those
driven by age or education in this example). From a policy standpoint we can
change the intervention but we cannot change the education or age of people, so
we want to remove the influence of age and education (among other relatively
fixed characteristics of people) from our description. That is, we want to be
able to **interpret** the relationship we calculate between intervention and
outcome as reflecting, as much as possible, only the intervention and the
outcome and not age and education.^[We borrow the idea of "interpretable
comparisons" from @kinder1993behalf]

# The simplest matched design: 2 strata, 1 variable

As an example of where we are going, we show an example of what is, in essence,
a matched design, but one that has only 2 strata and using only 1 variable.
Here we will also show the outcome analysis, just to be clear about how
stratification can enable adjustment or substitute for linear additive
"controlling for".

Say that we want to remove the effect of sex from the relationship between the
intervention and the outcome. The simplest and most transparent way to do this
would be to calculate the relationship between the intervention and the outcome
only among men and only among women (imagining that our administrative data do
not yet record non-binary gender).

Below we show this most basic approach to using stratification to remove or
control for the effects of a single variable with two levels as a way to
introduce multivariate optimal matching later. What we see here is that the
effect among men is different from the effect among women, and that the effect
among each sex cannot be confounded by sex --- after all the variables `sex` is
held perfectly constant.

```{r two_strata_example}
with(dat, table(treated=trtbin, sex=sex, exclude = c()))
lm_sexM <- lm(postY ~ trtbin, data = dat, subset = sex == "M")
lm_sexF <- lm(postY ~ trtbin, data = dat, subset = sex == "F")
coef(lm_sexM)[["trtbin"]]
coef(lm_sexF)[["trtbin"]]
```

Now, these are two descriptions of the relationship of interest, but we tend to
think about stratification as a way to *remove* confounding relationships from
*a single overall* relationship. How to calculate this? We show this in steps
here --- first doing it very simply by hand, and then using "fixed effects". We
will return to these ideas after creating matched designs using `optmatch` and
`designmatch` or another approach to stratification since this is the same way
that we will estimate effects after creating hundreds of matched pairs or sets
later.

We know how to analyze a block-randomized (or strata-randomized) experiment
[@gerber2012field]: each block is a mini-experiment. We *estimate the ATE
within each block* and *combine by weighting each block specific estimate*.

The *block-size weight* produces an unbiased estimator in randomized
experiments --- in an observational study we don't know about the bias since we
don't exactly know how to repeat the study.  The *precision weight* (aka the
"fixed effects" weights) tends to produce smaller standard errors and
confidence intervals but is biased in randomized experiments. In standard
practice with matched designs we use the *precision weight* approach because
(1) we tend to have matched sets that are nearly the same size (i.e. pairs)
which means that the block-size weight and the precision-weight approaches tend
to be nearly the same and (2) we can calculate the precision-weighted results
quickly using the fact that "fixed effects" in linear regression models create
precision weights and (3) we know that our estimators are biased because we
neither observe all relevant confounders nor do we know the functional form
that relates them to the treatment.

```{r weighting}
## First collapse the data to the level of the strata and calculate weights
dat_sets <- dat %>%
  group_by(sex) %>%
  summarize(
    nb = n(),
    ateb = mean(postY[trtbin == 1]) - mean(postY[trtbin == 0]),
    prob_trt = mean(trtbin), ## proportion treated in the strata
    nbwt = n() / nrow(dat), ## proportion of the total data in the strata
    prec_wt = nbwt * prob_trt * (1 - prob_trt),
  )

dat_sets$prec_wt_norm <- with(dat_sets, prec_wt / sum(prec_wt))

## The two groups of people with the treatment estimated within each set along with some weights
dat_sets %>% select(sex, nb, ateb, prob_trt, nbwt, prec_wt_norm)

## The overall estimation using the two kinds of weights
est_ate_nw <- with(dat_sets, sum(ateb * nbwt))
est_ate_pw <- with(dat_sets, sum(ateb * prec_wt_norm))

c(block_size_wt = est_ate_nw, precision_wt = est_ate_pw)
```

Strata-level weights can also be represented at the individual level --- and
this allows us to use linear models (least squares) to produce block-weighted
estimates of the overall average causal effect after "holding constant" sex.

```{r make_strata_level_data}
## Now creating the weights at the individual level
dat <- dat %>%
  group_by(sex) %>%
  mutate(
    nb = n(),
    mb = sum(trtbin),
    ateb = mean(postY[trtbin == 1]) - mean(postY[trtbin == 0]),
    prob_trt = mean(trtbin),
    nbwt = (trtbin / prob_trt) + (1 - trtbin) / (1 - prob_trt),
    prec_wt = nbwt * prob_trt * (1 - prob_trt)
  ) %>%
  ungroup()

## Two ways to use the block-size weight
est_ate1a <- difference_in_means(postY ~ trtbin, blocks = sex, data = dat)
est_ate1b <- lm_robust(postY ~ trtbin, weights = nbwt, data = dat)
est_ate1c <- lm(postY ~ trtbin, weights = nbwt, data = dat)

## Three other ways to use the precision or harmonic weight
est_ate2a <- lm_robust(postY ~ trtbin + sex, data = dat)
est_ate2b <- lm_robust(postY ~ trtbin, fixed_effects = ~sex, data = dat)
est_ate2c <- lm_robust(postY ~ trtbin, weights = prec_wt, data = dat)
```

Notice that we get the same result for the overall effect whether we first
calculate differences of means within strata and then take a weighted average
of those strata specific differences of means, or whether we hand off that job
to least squares. The only difference here is whether we choose to weight only
using the proportion of the data in the block, or whether we also want to take
into account the proportion taking the intervention within the strata.

```{r}
## Block-size weighted results
c(est_ate_nw, coef(est_ate1a)[["trtbin"]], coef(est_ate1b)[["trtbin"]], coef(est_ate1c)[["trtbin"]])
## Precision weighted results
c(est_ate_pw, coef(est_ate2a)[["trtbin"]], coef(est_ate2b)[["trtbin"]], coef(est_ate2c)[["trtbin"]])
```

## Summary of the section

So, we have seen that we can "control for" a single binary variable using
stratification. We didn't need to evaluate the stratification --- after all, we
just held sex perfectly constant within strata. If someone asked, "Does this
design remove sex as an alternative explanation?" we could answer simply, "Yes.
We only compare women with women and men with men." And we were able to produce
an estimate of an overall effect by calculating effects within strata and
combining them via weighting. And we showed that we can use OLS for this
purpose --- just as we would if we had a block-randomized experiment.

What should we do if we have more than one variable? Or if those variables have
more than a few values? We explain that below as we introduce optimal matching.

# Create a design, evaluate and iterate if needed

When we created our fake data we made two intervention variables, `trtbin` and
`trtcont`, because we want to demonstrate how to use two kinds of approaches to
stratification here: **bipartite matching** (which creates sets or pairs of people
who differ in a binary variable (i.e. comparing those who received the
intervention versus did not) and **non-bipartite matching** (which creates pairs of
people who differ in levels of a variable that has more than two values such
that each pair should have one person with a higher value and another person
with a lower value).

Imagine that we want to compare people who are similar in multiple demographic
characteristics as well as baseline outcome (`preY`). We would like to **create
a research design** where people are placed into strata where they are as
similar as possible with other people along all of those characteristics. If we
calculate effects within those strata, we know that we will have minimized the
effects of those characteristics on the outcome even if we have not exactly
eliminated them as we did above when we exactly controlled "male" versus
"female". In the last decades, algorithms to create strata that minimize
differences between people on those covariates have been implemented in
relatively user-friendly software (see the list in [@rosenbaum2020modern]). In
this guide we demonstrate the use of `optmatch` and `designmatch` because each
uses a slightly different approach to the minimization problem (both are
"optimal" because they solve problems by minimizing).

We demonstrate here how to do this by (1) combining the many variables into
single scores via *dimension reduction* and then using the optimal matching
algorithms to find strata that minimize the within-strata distances between
people on that score (say, using a Mahalanobis distance score and/or a
propensity score) and (2) by directly constraining the algorithm (say,
restricting research designs to only combine people with fewer than 10 years
difference in age).

## Bipartite Matching with Optmatch

We will start with the task of comparing people based on their experience of a
binary intervention (`trtbin`) using the `optmatch` package.

The general idea in optimal matching is to answer the question, "How shall we
break the data into strata?" with "We should choose strata within which
differences between units are minimized." Current software like `optmatch` and
`designmatch` create these strata via minimization using **distance matrices**.
That is, if we can represent differences between all of the treated and control
observations as some form of distances, then we can ask those optimal matching
algorithms to find strata that are optimal in the sense of there being no other
sets of strata where treated and control units are closer together on that
distance. For example, below we make a matrix of absolute differences between
each treated unit and each control unit in `preY`:

```{r abs_dist}
## Scalar/Absolute differences in baseline outcome
abs_dist <- match_on(trtbin ~ preY, data = dat, method = "euclidean")
abs_dist[1:3, 1:4]
## Notice that the distance between the treated unit labeled "1" and the control unit labeled "5" in the matrix above is simply the absolute difference:
abs(dat["1", "preY"] - dat["5", "preY"])
## Optmatch offers us some overview of this distance matrix
summary(abs_dist)
```

## Dimension reduction

We convert many columns into one column we reduce the dimensions of the dataset
(to one column). We can use the idea of **multivariate distance** to produce
distance matrices to minimize **multivariate distances**.

We defer in-depth explanation of these scores to [@rosenbaum2020book] for now.
I already created the distance matrix recording absolute differences on a
single variable `abs_dist` which records the absolute difference in the
pre-outcome between people in the treated versus control groups. Below we
create multivariate distances in two different ways: a Mahalanobis distance
(`mh_dist`) and a standardized distance on propensity score (`ps_dist`). The
Mahalanobis distance matrix records the differences in Mahalanobis distance
(after transforming the covariates to ranks and scaling, see the
[@rosenbaum2020book] section 9.3 on this); and `ps_dist` which records
differences in predicted treatment or "propensity score" on the linear
predictor scale after standardizing.

This next code chunk creates a rank-based **Mahalanobis distance matrix** using 9 covariates.

```{r mh_dists}
### Setup for multivariate distances
covs <- c("sso", "age", "sex", "race", "edu", "grade", "tenure", "preY")
cov_fmla <- reformulate(covs, response = "trtbin")
## A multivariate distance score: Mahalanobis distance on rank transformed covariates
mh_dist <- match_on(cov_fmla, data = dat, method = "rank_mahalanobis")
mh_dist[1:3, 1:4]
## person 213 is closer to person 1 in multivariate terms than person 104
mh_dist[1, c("104", "213")]
## Here we can compare those two people with person 1 along all of the covariates:
dat %>%
  filter(id %in% c(1, 104, 213)) %>%
  select(one_of(c("trtbin", "id", covs)))
summary(mh_dist)
```

This next produces a different distance matrix. Whereas the Mahalanobis
distance aims to weight each covariate equally, the **propensity distance** weights
the influence of each covariate in the score depending on how predictive it is
of the treatment. We often use a logistic regression to produce these scores,
but logistic regressions often overfit the data, so below we use a version of
logistic regression with the coefficients shrunken towards zero. What really
matters here is that the covariates that are *relatively more predictive* than
others have larger coefficients in absolute value.

```{r ps_dist}
## A multivariate distance score: Propensity score using a logistic regression
## model that shrinks coefficients toward zero to avoid the separation problem
## common in logistic models
## Not loading the arm package because it contains lots of stuff that we don't need so just
## using arm::bayesglm

psmod <- arm::bayesglm(cov_fmla, data = dat, family = binomial())
## Inspect the coefs to make sure that none are too huge
zapsmall(coef(psmod))
ps_dist <- match_on(psmod, data = dat)
ps_dist[1:3, 1:4]
ps_dist[1, c("104", "169")]

## Here we can compare those two people with person 1 along all of the
## covariates.
## Notice that control person 169 is closer to treated person 1
## across nearly all of the covariates than person 104 in regards to the
## propensity score

dat %>%
  filter(id %in% c(1, 104, 169)) %>%
  select(one_of(c("trtbin", "id", covs)))
summary(ps_dist)
```

## Creating the stratification using the `optmatch` package

Creating the stratification using the `pairmatch` and `fullmatch` commands in
the optmatch package just requires that the researcher provide an appropriately
formatted distance matrix to those functions.

Here is a **paired design**. The `pairmatch` command creates an object that is a
factor variable that indicates which individual belongs with which pair. Below
we see persons 1 and 3 (people who experienced the intervention) were placed
into pairs with person 627 and 383 (people who did not experience the
intervention). These people don't look extremely similar, so we may want to
fine tune this design using other information.

```{r pm1}
## A paired design
pm1 <- pairmatch(mh_dist, data = dat)
summary(pm1)

stopifnot(all.equal(names(pm1), row.names(dat)))
dat$pm1 <- factor(pm1)

dat %>%
  filter(pm1 %in% c("1.1", "1.2")) %>%
  select(one_of(c("pm1", "trtbin", "id", covs))) %>%
  arrange(pm1, trtbin)
```

We can also ask for a match with a variable number of controls and treated
units in each set where the number of controls per treated and treated per
control are chosen to minimize overall balance using **fullmatch**
[@hansen:2004]. Below we show a fully-matched design called  `fm1_unrestricted`
in which no control units are excluded. And we compare this design to one where
we restrict this design to have no more than 1 treated unit per set and no more
than 5 controls per set.^[We can make other decisions to generate a design that
we like later: the decision to restrict the size of the sets has mostly to do
with the fact the effective sample size of a stratified research design with
two treatments is enhanced as we (a) get more strata and (b) the strata are
more equal in size of treated versus controls. We can explain a lot more about
this.]

```{r fm1}
fm_unrestricted <- fullmatch(mh_dist, data = dat)
summary(fm_unrestricted, min.controls = 0, max.controls = Inf)

stopifnot(all.equal(names(fm_unrestricted), row.names(dat)))
dat$fm_unrestricted <- factor(fm_unrestricted)

dat %>%
  filter(fm_unrestricted %in% c("1.1", "1.167")) %>%
  select(one_of(c("fm_unrestricted", "trtbin", "id", covs))) %>%
  arrange(fm_unrestricted, trtbin)

## A full-matched design with no more than one treated unit per set and no more than 5 controls per set
fm1 <- fullmatch(mh_dist, min.controls = 1, max.controls = 5, data = dat)
summary(fm1, min.controls = 0, max.controls = Inf)

stopifnot(all.equal(names(fm1), row.names(dat)))
dat$fm1 <- factor(fm1)

dat %>%
  filter(fm1 %in% c("1.1", "1.10")) %>%
  select(one_of(c("fm1", "trtbin", "id", covs))) %>%
  arrange(fm1, trtbin)
```

In each case the strata (pairs or matched sets) are factor variables named with levels like
"1.1" or "1.99" etc.. where these names are not numbers, but "1.1" means "group
1, first set". And in our case, here we only have one overall group so all of
the sets start with `1`: when we use exact matching the first element of
the set name will change. We will show this below.

## How to assess a matched design?

Did either of those designs do a good job? Should we feel confident using them
to remove the influence of the covariates? We have two ways to answer those
questions.^[Another question we should ask about a research design is whether
it provides enough information to help us discern differences should those
differences exist. We set aside this question of statistical power or
information for later. From the perspective of stratified designs, statistical
power or information or "effective sample size" is a function of not only
number of total observations, but of the number of strata, and the variance of
both the treatment and the outcomes within strata. In the case of a binary
treatment, variance of the treatment, amounts to the proportion treated.  In a
matched design that creates pairs, these considerations go away --- the number
of strata is maximized and each pair has exactly one treated and one control
unit. When creating full matched designs --- with varying numbers of treated
and controls per set --- these issues become more important.]

### Look for substantively large strata-wise differences

First, we can answer this question about design adequacy by referring to what
we know about the units and the theory of change and the alternative
explanations that we are trying to engage. For example, imagine that we really
must remove the effect of age from our comparisons in order for any reasonable
interpretation of our analysis to shed light on the effect of the intervention
itself and not on age. We might then calculate the difference in age within
each set and ask whether any set has such a large difference in age that we
could not credibly claim to have "adjusteded away" the potentially confounsing
effects of age with this design overall.


```{r echo=FALSE, out.width=".9\\textwidth"}
library(gridExtra)
dat_plot1 <- dat %>% select(age,trtbin,fm1) %>%
  group_by(fm1) %>%
  mutate(agediff=abs(mean(age[trtbin==1]) - mean(age[trtbin==0])),
         nb=n()) %>%
  ungroup() %>%
  mutate(fm1 = factor(fm1, levels = names(sort(tapply(agediff, fm1, mean),decreasing=TRUE))))
dat_plot1$nostrata <- rep(1, nrow(dat_plot1))
```

For example, we could make a plot to show this information within sets versus
raw differences from the fullmatch. @fig-fm_eval_age shows, for example, we
show the worst 100 sets out of the 334 total sets in order of remaining absolute
difference in mean age between the treated and control observations from left
to right. Notice that no set has nearly the age difference that we see overall
(in the boxplot at the far right): without matching we would be comparing
people who are `r min(dat$age)` with people who are `r max(dat$age)` whereas
after matching the worse set has has a difference of `r
max(dat_plot1$agediff)`.

```{r}
#| echo: false
#| output: true
#| label: fig-fm_eval_age
#| fig-cap: "Within set absolute differences in mean age shown within each set (left panel) and overall, before any stratification (right panel). Red dots are the means of the absolute differences."
#| out-width: 75%
bpfm1 <- ggplot(dat_plot1 %>% filter(agediff > quantile(agediff,.9)), aes(x = fm1, y = age)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red") +
  geom_text(aes(x=fm1,y=min(age),label=nb))+
  annotate("text",x = 0, y = min(dat_plot1$age), label="Set \n size")
bpfm1 <- bpfm1 + scale_x_discrete(expand = expand_scale(add = c(2, 0.5)))

bporig <- ggplot(dat_plot1, aes(x = nostrata, y = age)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, geom = "point",
    shape = 20, size = 3, color = "red", fill = "red"
  )

grid.arrange(bpfm1, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

We can futher reason about these differences by calculating their quantiles.
For example, below we see that the worst matched set in the `fm1` design has a
difference of 33 years --- including a 15 year old and a 49 year old. This
might be a bad set depending on the context. Across the sets, we see that half
of them differ by less than 5.8 years of age, and that 90% of the sets differ
by less than about 13 years of age. Is this sign that we have done enough to
remove the effects of age from the impact evaluation? That depends on context.
Let us imagine for now that we are not satified with this design --- that we'd
like to have smaller differences on age. But that we also are worried about
some of the other variables. And, in fact we'd like to evaluate the design as a
whole. How might we do that?

```{r sdiffs, echo=FALSE}
rawmndiffs <- with(dat, mean(age[trtbin == 1]) - mean(age[trtbin == 0]))
setdiffsfm1 <- dat %>%
  group_by(fm1) %>%
  summarize(
    mn_age_diffs =
      mean(age[trtbin == 1]) - mean(age[trtbin == 0]),
    mn_age= mean(age),
    min_age= min(age),
    max_age= max(age)
  )

setdiffsfm1 %>% arrange(desc(abs(mn_age_diffs))) %>% head()
## summary(setdiffsfm1$mn_age_diffs)
quantile(abs(setdiffsfm1$mn_age_diffs), seq(0,1,.1))
```

### How to assess stratified research designs as a whole: the omnibus-testing approach

To evaluate the design as a whole, let us start with what we know. Consider a
standard against which to compare our design, the block- or strata-randomized
experiment. We know that in designs randomized within strata:

 - Randomization balances covariate distributions between treated and control
 groups. **It does not make them identical.**
- We can repeat the known randomization to check the randomization procedure
(mostly useful if there is a long chain of communication between the random
number generator and the field such that we worry about the implementation of
the experiment rather than the random number generator on our computer). This
repetition is useful because **we know how strata weighted mean differences in
covariates would vary in an experiment.**
- **Randomization does not imply exact equivalence. Large differences in
covariates easily arise in small experiments.**

This means that we can compare  **stratified observational studies** to
equivalent randomized experiments [@hansen2008cbs].

The within-set differences look different from those that would be expected
from a randomized experiment.^[If you look at the documentation for
`balanceTest` you'll that it can also handle clustered data as well as other
options.] Below we see that (1) the un-adjusted design (represented with `--`
in the `overall` comparison of the data to the standardized mean differences in
covariates we would have expected from a randomized experiment) does not look
consistent with the experimental standard (the $p$-value is so tiny that we
would say that nearly no randomized experiments would have the pattern of
imbalance across all of those covariates that we see in our data); (2) And that
while the full-matching (`fm1`) and  pair-matching (`pm1`) improve the
imbalance slightly, neither looks appreciably consistent with a randomized
experiment (again, the $p$-values are too small). If, however, we saw a
$p$-value of, say, $p=.8$ we would say that the given design has smaller
covariate imbalance than 80% of equivalent randomized experiments --- which
would be a sign that our design compares favorably with randomized experiments.

```{r compare_bal}
#| eval: true
#| echo: true

balfmla_new <- stats::update(cov_fmla,.~.+ strata(fm1)+strata(pm1))
balfmla_new
# trtbin ~ sso + age + sex + race + edu + grade + tenure + preY +
#     strata(fm1) + strata(pm1)

xbfm1 <- balanceTest(balfmla_new, data = dat, p.adjust.method="holm")
xbfm1$overall
```

Notice that you can also use `independence_test` from the `coin` package for
this purpose. This function requires a bit more work on the part of the user,
but we show it here just to be clear that this approach can be implemented via
at least two pre-packaged pieces of software.

```{r coin_omni_demo}

## coin doesn't know what to do with all of these variables of mixed types plus
## it wants a factor variable as the treatment, so the model.matrix command
## expands all of the factors into 0,1 variables. balanceTest does this behind
## the scenes

dat$trtbinF <- factor(dat$trtbin)

##datmm <- data.frame(stats::model.matrix(stats::update(balfmla_new,~.-1-strata(fm1)-strata(pm1)),data=dat))
datmm <- fill.NAs(stats::update(balfmla_new,~.-1-strata(fm1)-strata(pm1)),data=dat,all.covs=TRUE)

datmm$trtbinF <- dat$trtbinF
datmm$trtbin <- dat$trtbin
datmm$fm1 <- dat$fm1

balfmla_new_for_comparison <- reformulate(names(datmm)[!(names(datmm) %in% c("trtbin","trtbinF","fm1"))],response="trtbin")
## trtbin ~ sso + age + sexF + sexM + raceB + raceW + edub + educ + edud + edue + eduf + edug + eduh + grade + tenure + preY

## The version using asymptotic approximations to do the randomization inference
bt_fm1 <- balanceTest(stats::update(balfmla_new_for_comparison,.~.-1+strata(fm1)),data=datmm)
bt_fm1$overall["fm1",]
xb_fm1 <- xBalance(balfmla_new_for_comparison,strata=list(fm1=~fm1),data=datmm,report="all")
xb_fm1$overall["fm1",]

## Now independence_test wants the formula to be flipped (rhs versus lhs)
library(formula.tools)
coin_fmla <- ~ trtbinF | fm1

lhs(coin_fmla) <- formula.tools::rhs(balfmla_new_for_comparison)
coin_fmla
## The version using asymptotic approximations to do the randomization inference
d2_coin_test_asymp <- independence_test(coin_fmla,data=datmm,teststat="quadratic",distribution=asymptotic())
d2_coin_test_asymp
pvalue(d2_coin_test_asymp)

rbind(c(statistic(d2_coin_test_asymp),pvalue(d2_coin_test_asymp)),
bt_fm1$overall["fm1",c(1,3)],
xb_fm1$overall["fm1",c(1,3)])

```

We can also inspect the individual mean differences across many covariates in
one table (shown below). We have sorted the table by the variables that have
the largest $z$ statistics (calculated using a large-sample approximation to
the randomization-based distribution of the null hypothesis of no differences
on the covariate between treated and control groups conditional on strata). The
$p$-values are adjusted for multiple testing using the Holm-Bonferroni method
since we do not want to mislead ourselves by looking at 19 $p$-values and
declaring any $p < .05$ as "significantly imbalanced". We see that there are
still appreciable differences in baseline outcome (`preY`) and `age` (a
difference of about 5 years after adjusting for the stratification) and  in the
proportion of White respondents (a difference of about 6 percentage points).

```{r}
xbfm1_dat <- data.frame(xbfm1$results[,,"fm1"])
xbfm1_dat %>% arrange(desc(abs(z)))
```

We will return to the question of how to make the balance better below when we
talk about exact matching and the use of calipers in the `optmatch` package and
the use of "nearness" penalties in the `designmatch` package. First, however,
we will explain a bit more about what `balanceTest` is doing --- it turns out
that the general approach for assessing balance is the same as the approach we
will use to estimate average treatment effects once we are satisfied with the
design.


### What is balanceTest doing?

The balanceTest approach compares the strata-weighted average of within-strata differences to that
which would be expected if we were to repeat an experiment with the same
stratified design and same covariate values (and `balanceTest` uses a large
sample Normal approximation to this distribution.)

```{r xbagain, echo=TRUE}
setmeanDiffs <- dat %>%
  group_by(fm1) %>%
  summarise(
    diffage = mean(age[trtbin == 1]) - mean(age[trtbin == 0]),
    nb = n(),
    nTb = sum(trtbin),
    nCb = sum(1 - trtbin),
    hwt = (2 * (nCb * nTb) / (nTb + nCb))
  )
setmeanDiffs
```


The test statistic is a weighted average of the set-specific differences (*same
approach as we would use to test the null in a block-randomized experiment*).
What weight do we use? In this case, `balanceTest` reports as the **descriptive**
strata-adjusted mean difference the average of the mean differences within each
strata weighted by the number of treated in each strata, one might call this
the average effect of treatment on the treated style weighting.

```{r wtmns, echo=TRUE}
## The descriptive mean difference using block-size weights
with(setmeanDiffs, sum(diffage * nTb / sum(nTb)))
## Compare to balanceTest output
xbfm1$results["age","adj.diff" , "fm1"]
```

However, the **test statistic** used to create the $p$-values uses a different
weighting scheme which is known to produce the most powerful tests of the null
of no differences [@hansen2008cbs]. Hansen and Bowers call this a harmonic-mean
weight. We could also call it a precision-weight. Or a "fixed-effects" weight.
This weight considered both the overall size of the set but also the ratio of
treated to control units within each set (both of which would determine the
standard error of the estimate of the average treatment effect if it were
calculated within a given set).

```{r}
## The mean diff used as the observed value in the testing is weighted
with(setmeanDiffs, sum(diffage * hwt / sum(hwt)))
coef(lm_robust(age~trtbin,fixed_effects=~fm1,data=dat))
```

#### Where does the null distribution for the overall test statistic come from?

We will show this distribution in detail here --- generating it by hand using
code that should be relatively easy to understand (compared to the matrix
algebra used for efficiency within the RItools package). But we can also
calculate this $p$-value directly using permutations of treatment and control
within set using the `coin` package as shown above:

```{r}

d2_coin_test_asymp <- independence_test(coin_fmla,data=datmm,teststat="quadratic",distribution=asymptotic())
d2_coin_test_asymp
pvalue(d2_coin_test_asymp)

rbind(c(statistic(d2_coin_test_asymp),pvalue(d2_coin_test_asymp)),
bt_fm1$overall["fm1",c(1,3)],
xb_fm1$overall["fm1",c(1,3)])

## Notice that the balanceTest from RItools and independence_test from coin have the same test statistic (the precision weighted version)
#cbind(xb_fm1$results[,c("adj.diff","std.diff","z"),],coin_test_stat=matrix(statistic(d2_coin_test_asymp,"standardized"),ncol=1),
#bt_fm1$results[,c("adj.diff","std.diff","z"),])
cbind(bt_fm1$results[,c("adj.diff","std.diff","z"),],coin_test_stat=matrix(statistic(d2_coin_test_asymp,"standardized"),ncol=1))

## The version using permutations to do the randomization inference
d2_coin_test_perm<- independence_test(coin_fmla,data=datmm,teststat="quadratic",distribution=approximate(nresample = 10000))
d2_coin_test_perm

## Draw from the two different approximations to the null distribution from the two packages
## Using the same seed for each drawing approach
set.seed(12345)
## These two use the coin package rperm function to draw from the distributions implied by the independence_test (a chisq in one case, and a permutation dist in the other)
null_perm_dist <- rperm(d2_coin_test_perm,10000)
set.seed(12345)
null_asymp_dist <- rperm(d2_coin_test_asymp,10000)
## This next is what is used by balanceTest
set.seed(12345)
bt_dist <- rchisq(n=10000,df=16)
pchisq(bt_fm1$overall["fm1",1],15,lower.tail=FALSE)
bt_fm1$overall["fm1",]

d2_dists <- data.frame(tstat=c(null_perm_dist,null_asymp_dist),type=rep(c("perm","asymp"),each=10000))

dist_compare <- ggplot(d2_dists,aes(x=tstat,group=type,color=type))+
  geom_density()

dist_compare
```

Notice that the $\chi^2$ distributions are very closely approximating the permutation based distribution?

Now we can do the permutations by hand to be more clear about what is going on.
We first calculate the $d$-statistics (which are just stratum-weighted
differences in means), and then $d^2$ statistic. We generate their
distributions using permutations that repeat the design of the
putative/metaphorical randomized experiment.

```{r}

d_stat <- function(covs_dat,trt,strat){
  dat <- cbind(covs_dat,z=trt,s=strat)
  cov_nms <- names(covs_dat)
  datb <- dat %>% group_by(s) %>%
    summarize(across(.cols=one_of(cov_nms),function(x){ mean(x[z==1]) - mean(x[z==0])}),
  nb=n(),
  pib=mean(z),
  nbwt=nb/nrow(dat),
  hbwt0= pib * (1-pib) * nbwt) %>% ungroup()
  datb$hbwt <- datb$hbwt0/sum(datb$hbwt0)
  adjmns <- datb %>% summarize(across(.cols=all_of(cov_nms),function(x){ sum(x*hbwt) }))
  adjmnsmat <- as.matrix(adjmns)
  return(adjmnsmat)
}

test_d_stat <- d_stat(covs_dat=dplyr::select(datmm,-one_of(c("trtbinF","trtbin","fm1"))),trt=datmm$trtbin,strat=datmm$fm1)
## xb_fm1$results[,"adj.diff","fm1"] ## for comparison, the older xBalance function reports the inferential test stats and not the descriptive ones

d2_stat <- function(dstats,ddist=NULL,theinvcov=NULL){
  ## d is the vector of d statistics
  ## ddist is the matrix of the null reference distributions of the d statistics
  if(is.null(theinvcov) & !is.null(ddist)){
    as.numeric( t(dstats) %*% solve(cov(t(ddist))) %*% dstats)
  } else {
    as.numeric( t(dstats) %*% theinvcov %*% dstats)
  }
}

## Now generate the distribution of the d statistics under the null
new_randomization <- function(z,s){
  ## This is for conceptual understanding. So it will be slow.
  newz <- data.frame(z,s) %>% group_by(s) %>% mutate(newz=sample(z)) %>% ungroup() %>% dplyr::select(newz) %>% unlist()
  return(newz)
}

## This should keep the same number of treated and controls in each strata as the original
## we are just shuffling treatment within strata in the same way that we would imagine to
## have done so if we were assigning treatment in a block randomized trial

test_new_rand <- new_randomization(datmm$trtbin,datmm$fm1)
stopifnot(all.equal(table(z=datmm$trtbin,s=datmm$fm1),table(z=test_new_rand,s=datmm$fm1)))
cov_nms <- names(dplyr::select(datmm,-one_of(c("trtbinF","trtbin","fm1"))))

#d_stat(trt=new_randomization(datmm$trtbin,datmm$fm1), covs_dat=datmm[,cov_nms], strat= datmm$fm1)

set.seed(12345)
## This next is fairly slow
d_dist<-replicate(10000, d_stat(trt=new_randomization(datmm$trtbin,datmm$fm1),
  covs_dat=datmm[,cov_nms],
  strat= datmm$fm1))
## only for now:
d_dist <- drop(d_dist) ## for some reason d_dist comes out an array
d_dist[1:3,1:5]

```

```{r}
#| eval: true

d_dist_cov_mat <- cov(t(d_dist))
inv_dist_cov_mat <- MASS::ginv(d_dist_cov_mat)
## We use the generalized inverse because the cov matrix is nearly singular (not exactly though).
## So, the following produces an error
## invCovDDist <- solve(cov(t(d_dist)))
# Here we can see the preY is close to a linear combination of the other columns
##eigen(d_dist_cov_mat)

## The observed values of the test statistics:
obs_d <- drop(d_stat(trt=datmm$trtbin,covs_dat=datmm[,cov_nms],strat=datmm$fm1))
obs_d2<- d2_stat(obs_d,d_dist,inv_dist_cov_mat)

## The distribution of the d2 test statistic
d2_dist<-apply(d_dist, 2, function(thed){
                 d2_stat(thed,theinvcov=inv_dist_cov_mat)
         })

## The chi-squared reference distribution only uses a one-sided p-value going in the positive direction
d2_p<-mean(d2_dist>=obs_d2)
cbind(obs_d2,d2_p)
bt_fm1$overall

length(d2_dist)

d2_dists <- data.frame(tstat=c(null_perm_dist,null_asymp_dist,d2_dist),type=rep(c("perm","asymp","perm_by_hand"),each=10000))

dist_compare <- ggplot(d2_dists,aes(x=tstat,group=type,color=type))+
  geom_density()

dist_compare

```

As the plot shows, all three approaches to the reference distribution --- direct permutation, the $\chi^2$ large-sample approximation, and our hand-built permutation of the $d^2$ statistic --- agree closely, confirming that `balanceTest` is doing exactly what we described above.

So, we can see that the omnibus test is a test the combines the tests of the
nulls for each individual variable. The use of the $\chi^2$ distribution as a
large sample approximation to the randomization distribution arises from the
fact that we taking a linear combination (a sum) of mean differences, each of
which is squared and standardized and has its own Normal distribution (in large
samples like the one we are working with here). We have also seen that we can
apply the $d^2$ and $d$ tests directly by repeatedly reshuffling the treatment
assignment following the matched-design within strata. This foreshadows our
recommended approach to estimating causal effects and testing hypotheses about
causal effects --- we will do so with reference to an as-if-randomized
treatment assignment process within each of the strata created by the matching
procedure.


## Using Designmatch for bipartite matching

Another tool for creating matched designs when you have one intervention group
and one control group is provided by the `bmatch` function in the `designmatch`
package. It has many options that allow fine-tuning of the stratified design.

To use this function the data have to be sorted in decreasing order of the treatment vector:

```{r}
library(designmatch)
library(highs)

datmm_dm <- datmm[order(datmm$trtbin,decreasing=TRUE),]

## Convert propensity distance matrix into a form useable by bmatch (because the data was re-ordered):
ps_dist_dm_0 <-as.matrix(ps_dist)[row.names(datmm_dm)[datmm_dm$trtbin==1],row.names(datmm_dm)[datmm_dm$trtbin==0]]
## designmatch is picky about how the distance matrix is scaled
ps_dist_dm<- round(ps_dist_dm_0/mean(ps_dist_dm_0),2)
## See the manual page for distmat() in the designmatch manual for its built in way to produce mahalanobis distance matrices

solverlist <- list(name = "highs", approximate = 1, t_max = 1000, trace = 1)

## bmatch does not do full matching. It defaults to pair matching and allows
## fixed numbers of controls (i.e. every match can be 1 to 2, or 1 to 3, but not a
## mixture of those)

## I'm still a little confused by the subset_weight argument. Sometimes I use
## the total_groups function instead. ## Here I am copying the call to distmatch()
## from designmatch where subset_weight=0 and the emphasis is on minimizing
## distances (just like the pairmatch function in optmatch)

dm1 <- bmatch(t_ind=datmm_dm$trtbin, dist_mat=ps_dist_dm, solver=solverlist, subset_weight=0, total_groups=sum(datmm_dm$trtbin),n_controls=1)

## there is no summary method or other methods to easily inspect the result,
## but we created a function to convert the results of bmatch to a factor like
## that produced by optmatch so that we can use tidyverse to make plots and
## describe quantiles of differences as well as use balanceTest and eventually
## estimate and test.

#' Convert a bmatch object to a data frame for use in merging onto an original data frame
#'
#' 
#' @param obj a bmatch object
#' @param origid the vector of identifications from the original data frame
#' @details This function has only been tested with pair matches
bmatch_to_dat <- function(obj,origid) {
  ## This is pair matching. Make sure of it.
  stopifnot(length(obj$t_id) == length(obj$c_id))
  ## We want a factor that we can merge onto our
  ## existing dataset. Here returning a data.frame so that
  ## we can merge --- seems less error prone than using
  ## rownames even if it is slower.
  matchesdat <- data.frame(
    set_id = obj$group_id,
    match_id = c(obj$t_id, obj$c_id)
  )
  matchesdat$id <- origid[matchesdat$match_id]
  return(matchesdat)
}

datmm_dm$id <- row.names(datmm_dm)
dm1_dat <- bmatch_to_dat(dm1,origid=datmm_dm$id)
datmm_dm_res <- left_join(datmm_dm,dm1_dat,by="id")
## This particular design dropped 332 people.
sum(is.na(datmm_dm_res$set_id))

```

We can evaluate the design produced by `bmatch` just as we did above. Here we just demonstrate balance test since the strata-quantiles approach code is the same as above (modified to use `set_id` rather than `fm1`)

```{r}
dm1_fm1 <- balanceTest(stats::update(balfmla_new_for_comparison,.~.-1+strata(set_id)),data=datmm_dm_res)
dm1_fm1$overall["set_id",]
```

### Tools to improve design

 So far we have not done a very good job in removing the relationships that
 exist between the covariates and the intervention (as can be seen from the
 omnibus balance test p-values or even from directly looking at each strata and
 the imbalance within them). So, we will try to improve the design by using a
 few different tactics: (1) Exactly matching based on some important
 categorical variables such as  `sex` (using either the `exactMatch()` function
 from `optmatch` or the `exact` argument to `bmatch`) (2) And requiring that
 members of strata differ by no more than a certain amount on either the
 dimension reduced score (like the propensity score) or one or more specific
 variables (using either the `caliper()` function from `optmatch` or the `near`
 argument to bmatch).

## Calipers

A **caliper** forbids pairs whose distance exceeds a threshold by setting those
entries in the distance matrix to `Inf`, making them unavailable to the
optimizer. This is useful both to eliminate the worst matches and to speed up
optimization on large datasets. In `optmatch` the `caliper()` function adds
these constraints to any distance matrix via ordinary addition.

### Caliper on a single variable

A common starting point is to inspect the distribution of distances and then
choose a caliper that rules out only the most extreme pairs. Here we apply a
caliper on the absolute difference in `preY` (baseline outcome), forbidding
pairs that differ by more than 0.08:

```{r caliper_one_dim}
## First inspect: how are the distances distributed?
quantile(as.vector(abs_dist), seq(0, 1, .1))

fm2 <- fullmatch(abs_dist + caliper(abs_dist, .08), data = dat)
summary(fm2, min.controls = 0, max.controls = Inf)

balfmla_new2 <- stats::update(balfmla_new,.~.+ strata(fm2))
xb2 <- balanceTest(balfmla_new2, data = dat, p.adjust.method="holm")

xb2$overall
xb2$results[, , ]
```

### Caliper on the propensity score

A very common practice is to apply a caliper of 0.2 propensity-score standard
deviations (roughly 2 standard deviations of the linear propensity score in many
applications). Here we first inspect the propensity-distance distribution and
then apply a caliper of 2 standard deviations:

```{r}
## Inspect the distance matrix itself
quantile(as.vector(ps_dist), seq(0, 1, .1))
## Apply a caliper (sets forbidden entries to Inf)
ps_distCal <- ps_dist + caliper(ps_dist, 2)
as.matrix(ps_dist)[5:10, 5:10]
as.matrix(ps_distCal)[5:10, 5:10]
summary(ps_distCal)
```

Now use the caliper-constrained propensity distance for full matching and pair
matching. The `propensity.model` option to `summary` gives a quick balance
check:

```{r}
fmCal1 <- fullmatch(ps_dist + caliper(ps_dist, 2), data = dat, tol = .00001)
summary(fmCal1, min.controls = 0, max.controls = Inf, propensity.model = psmod)
pmCal1 <- pairmatch(ps_dist + caliper(ps_dist, 2), data = dat, remove.unmatchables = TRUE)
summary(pmCal1, propensity.model = psmod)
```

### Multiple calipers

Calipers on different variables can be combined by adding them together. Here we
match on Mahalanobis distance but additionally disallow any pair that is too far
apart on the propensity score or on the baseline outcome:

```{r}
## Create a mahalanobis distance matrix (rank-transformed)
balfmla <- reformulate(c("tenure", "age"), response = "trtbin")
mhdist <- match_on(balfmla, data = dat, method = "rank_mahalanobis")

## Absolute difference in baseline outcome
tmppreY <- dat$preY
names(tmppreY) <- rownames(dat)
abs_dist <- match_on(tmppreY, z = dat$trtbin, data = dat)
abs_dist[1:3, 1:3]
quantile(as.vector(abs_dist), seq(0, 1, .1))
quantile(as.vector(mhdist), seq(0, 1, .1))

## Combine: match on MH distance, forbid pairs >9 MH units or >2 preY units apart
distCal <- ps_dist + caliper(mhdist, 9) + caliper(abs_dist, 2)
as.matrix(distCal)[5:10, 5:10]
as.matrix(mhdist)[5:10, 5:10]
```

Now create full matches from the combined-caliper distance. The `tol` argument
controls the optimizer's tolerance; a tighter tolerance yields smaller total
distances at the cost of longer run time:

```{r}
fmCal2a <- fullmatch(distCal, data = dat, tol = .001)
summary(fmCal2a, min.controls = 0, max.controls = Inf)
fmCal2b <- fullmatch(distCal, data = dat, tol = .00001)
summary(fmCal2b, min.controls = 0, max.controls = Inf, propensity.model = psmod)

dat$fmCal2a <- fmCal2a
dat$fmCal2b <- fmCal2b

fmCal2a_dists <- matched.distances(fmCal2a, distCal)
fmCal2b_dists <- matched.distances(fmCal2b, distCal)

mean(unlist(fmCal2a_dists))
mean(unlist(fmCal2b_dists))
```

As expected, the tighter tolerance (`fmCal2b`) achieves a lower mean within-stratum distance.

## Exact Matching

We often have covariates that are categorical/nominal and for which we really
care about strong balance. One approach to solve this problem is match
**exactly** on one or more of such covariates. If the process of matching is
going slow, this is also an approach to speed things up.

Here we demonstrate full matching but within categories defined by `grade`
(recoded to lo versus high) and gender:

```{r echo=FALSE}
dat$gradeLowHi <- ifelse(dat$grade %in% c(4,5,6), "hi", "lo")
with(dat,table(grade,gradeLowHi))
```

```{r}
dist2 <- ps_dist + exactMatch(trtbin ~ gradeLowHi+sex, data = dat)
summary(dist2)
## or mhdist <- match_on(balfmla,within=exactMatch(trtbin~gradeLowHi,data=dat),data=dat,method="rank_mahalanobis")
## or fmEx1 <- fullmatch(update(balfmla,.~.+strata(gradeLowHi)),data=dat,method="rank_mahalanobis")
fmEx1 <- fullmatch(dist2, data = dat, tol = .00001)
summary(fmEx1, min.controls = 0, max.controls = Inf, propensity.model = psmod)
print(fmEx1, grouped = T)
dat$fmEx1 <- fmEx1
```
```{r}
ftable(grade = dat$gradeLowHi, Trt = dat$trtbin, fmEx1, col.vars = c("grade", "Trt"))
```

### Exact matching with `designmatch`

The `bmatch` function supports exact matching through its `exact` argument,
which accepts a matrix of indicator variables --- one column per variable on
which exact matching is required. Every matched pair must share the same value
on every column.

```{r}
## Build the indicator matrix required by bmatch.
## Each column codes membership in one level of a categorical variable.
## bmatch will force every matched pair to match on all columns.
exact_vars <- model.matrix(~ gradeLowHi + sex - 1, data = datmm_dm)

dm_exact <- bmatch(
  t_ind    = datmm_dm$trtbin,
  dist_mat = ps_dist_dm,
  exact    = exact_vars,
  solver   = solverlist,
  subset_weight = 0,
  total_groups  = sum(datmm_dm$trtbin),
  n_controls    = 1
)

datmm_dm$id <- row.names(datmm_dm)
dm_exact_dat <- bmatch_to_dat(dm_exact, origid = datmm_dm$id)
datmm_dm_exact_res <- left_join(datmm_dm, dm_exact_dat, by = "id")

## How many treated units were left unmatched?
sum(is.na(datmm_dm_exact_res$set_id))

## Verify that every matched pair shares the same gradeLowHi and sex
matched_pairs <- datmm_dm_exact_res[!is.na(datmm_dm_exact_res$set_id), ]
ftable(gradeLowHi = matched_pairs$gradeLowHi,
       sex        = matched_pairs$sex,
       set_id     = matched_pairs$set_id)
```

Exact matching forces every stratum to be internally homogeneous on the
specified variables; the trade-off is that more units may be left unmatched when
the exact cells are sparse.

### Near-exact matching with `designmatch`

When strict exact matching would discard too many units, `designmatch` offers
**near-exact** constraints via the `near_exact` argument. Near-exact matching
penalizes pairs that differ on a categorical variable rather than forbidding
them outright. The penalty weight controls the trade-off between minimizing
covariate distance and matching exactly.

```{r}
## near_exact takes a list with element $covs (the matrix of categorical
## variables to penalize) and $devs (a scalar penalty weight per mismatch).
near_exact_vars <- model.matrix(~ gradeLowHi + sex - 1, data = datmm_dm)

dm_near_exact <- bmatch(
  t_ind    = datmm_dm$trtbin,
  dist_mat = ps_dist_dm,
  near_exact = list(covs = near_exact_vars, devs = 10),
  solver   = solverlist,
  subset_weight = 0,
  total_groups  = sum(datmm_dm$trtbin),
  n_controls    = 1
)

dm_near_exact_dat <- bmatch_to_dat(dm_near_exact, origid = datmm_dm$id)
datmm_dm_near_res <- left_join(datmm_dm, dm_near_exact_dat, by = "id")

## How many units were left unmatched?
sum(is.na(datmm_dm_near_res$set_id))

## How often did pairs differ on gradeLowHi or sex?
matched_near <- datmm_dm_near_res[!is.na(datmm_dm_near_res$set_id), ]
```

Near-exact matching typically retains more matched units than strict exact
matching while still strongly discouraging cross-category pairs. Increasing the
`devs` penalty pushes the solution toward exact matching; decreasing it toward
unrestricted distance minimization.

### A further note on exact matching: Splitting up the matching process by hand

If exact matching is still not working, one can directly split up the matching
process into separate pieces that would be defined by the exact matching calls
above.

In `optmatch` one can follow these kinds of ideas:
<https://cran.r-project.org/web/packages/optmatch/vignettes/matching-within-subgroups.html>.
For example, one can first produce the distance matrices of interest and then
use `findSubproblems()` to create a list of matrices, each of which can be
addressed as its own matching problem (in a loop, for example).

# Non-bipartite matching using `nmatch` from `designmatch`

# Estimating effects and testing hypotheses after matching

### After bipartite matching

### After non-bipartite matching

# Assessing sensitivity to unobserved confounders

```{r}
library(senstrat)
library(sensemakr)
```


# References


